Revisiting the Nos\'{e}-Hoover chain thermostat for Rigid-Body Molecular Dynamics

### 2018-02-04 rigid_bodies.tex
add figures
histograms
Change of format for the American Chemical Society

\documentclass[aip,jcp,reprint,amsmath,amssymb]{revtex4-1}

\usepackage[pdftex]{graphicx}
\usepackage[]{threeparttable}

\newcommand{\mt}[1]{\boldsymbol{\mathbf{#1}}}           % matrix symbol
\newcommand{\vt}[1]{\boldsymbol{\mathbf{#1}}}           % vector symbol
\newcommand{\tr}[1]{#1^\text{t}}                        % transposition
\newcommand{\diff}[2]{\dfrac{\partial #1}{\partial #2}} % partial derivative

\newcommand{\sn}{\text{sn}}                             % Jacobi function sn
\newcommand{\cn}{\text{cn}}                             % Jacobi function cn
\newcommand{\dn}{\text{dn}}                             % Jacobi function dn

\newenvironment{smallarray}[1]                          % small arrays
{\null\,\vcenter\bgroup\scriptsize
	\renewcommand{\arraystretch}{1.5}
	\arraycolsep=.13885em
	\hbox\bgroup$\left[\array{@{}#1@{}}}
{\endarray\right]$\egroup\egroup\,\null}

\title{Revisiting the Nos\'{e}-Hoover chain thermostat in the Rigid Bodies Molecular Dynamics}

@@

\date{\today}

%\begin{abstract}
%In Molecular Dynamics, sets of atoms collectively behaving as rigid bodies are often used to model entire molecules or parts thereof. Such a coarse-graining strategy eliminates degrees of freedom and admits larger time steps without abandoning the atomistic character of a model. In this paper, we rely on a particular factorization of the rotation matrix to simplify the mechanics of systems containing rigid molecules. Then, we present equations of motion and devise time-reversible, measure-preserving integrators for both microcanonical (NVE) and canonical (NVT) dynamics. We also develop new expressions for computing the virial pressure of rigid-body systems under periodic boundary conditions. Finally, simulations of liquid water are employed to analyze the numerical aspects of the proposed methodology.
%\end{abstract}

\maketitle

@@

Molecular Simulation, being the computational realization of Statistical Mechanics,\cite{Tuckerman2010} involves several types of approximations in order to mimic ``real'' physical systems. Naively, one would expect those approximations to only produce quantitative deviation from experimental values, and usually we tend to attribute any disagreement to a lack of accuracy of the force field. However, what can be hidden in the deviations are actually systematic errors, which may introduce non-physical behaviour. For instance, it has already been shown that the unavoidable truncation errors, associated with the numerical integration in Molecular Dynamics (MD), disrupt the equipartition of kinetic energy and, as a consequence, the temperature becomes an undefined property.\cite{Eastwood_2010} In a previous paper,\citep{Abreu_2017} we reported the same artifact to occur on the partition of kinetic energy between translational and rotational degrees of freedom in rigid-body MD. That finding brings into question the method introduced by Kamberaj and co-workers\cite{Kamberaj2005} which, assuming equipartition, couples separate chains of Nos\'{e}-Hoover thermostats on the rotational and translational degrees of freedom, but fails to reproduce the specified temperature for certain time step sizes.

@@

In this paper, we derive time reversible and measure-preserving numerical integrators for rigid-body MD in the canonical (NVT) ensemble. The starting point is our previous work on the microcanonical (NVE) ensemble,\citep{Abreu_2017} which employs the unit quaternion representation for the rotational degrees of freedom. Simultaneosuly coupling the translational and rotational degrees of freedom to a unique Nos\'{e}-Hoover chain thermostat, we are able to mantain the temperature of the system at its specified value. The analysis of the integrators includes comparisons with results from the method introduced by Kamberaj~\textit{et al.},\cite{Kamberaj2005} as well as with the Hybrid Monte Carlo (HMC) method,\cite{Duane1987} which is exact in the sense that truncation errors do not affect its accuracy. We also include results from simulations using the conventional SHAKE\cite{Ryckaert1977} procedure, also coupled to a unique Nos\'{e}-Hoover chain.

@@

Prompted by a particular factorization of the rotational matrix, in a previous paper\cite{Abreu_2017} we derived symplectic numerical integrators in the NVE ensemble: one that uses the exact solution for torque-free rotation and another approach, based on the method introduced by Miller~\textit{et al.},\cite{Miller2002}, that employs an approximate solution for rotation. For mathematical details the reader is referred to our paper.\cite{Abreu_2017} Next, we restrict ourselves to present the expressions required to develop the NVT case.

@@

The matrix $\mt A$ which can be computed from a given unit quaternion $\vt q$ as explained in Refs.~\onlinecite{Allen1989,Miller2002} admits the following factorization:
\begin{equation}
\label{eq:factorization_of_A}
{\mt A} = \tr{\mt B}{\mt C},
\end{equation} 
where
\begin{equation}
\label{eq:def_B_and_C}
\mt B(\vt q) = \begin{smallarray}{rrrr}
-q_2 & -q_3 & -q_4 \\
 q_1 & -q_4 &  q_3 \\
 q_4 &  q_1 & -q_2 \\
-q_3 &  q_2 &  q_1
\end{smallarray}
\; \text{and} \;
\mt C(\vt q) = \begin{smallarray}{rrrr}
-q_2 & -q_3 & -q_4 \\
 q_1 &  q_4 & -q_3 \\
-q_4 &  q_1 &  q_2 \\
 q_3 & -q_2 &  q_1
\end{smallarray}.
\end{equation}
That factorization of matrix $\mt A$ provides the mathematical relation between the three-dimensional angular velocity vector $\vt \omega$ and the four-dimensional rate of change of the quaternion components $\dot{\vt q}$, which is given by
\begin{equation}
\label{eq:relation_qdot_omega}
\dot{\vt q} = \frac{1}{2} \mt B \vt \omega.
\end{equation}
For the sake of brevity, the argument of a matrix-valued function (like $\mt A$, $\vt B$, or $\mt C$) might be omitted, when such argument is $\vt q$.

In the Hamiltonian approach, we employ the linear momentum $\vt p = m \dot{\vt r}$, conjugated to $\vt r$, as well as the momentum $\vt \pi= 4 \mt B \mt I \tr{\mt B} \dot{\vt q} = 2 \mt B \mt I \vt \omega$, conjugated to $\vt q$\cite{Goldstein2002}. In these equations, $m$ is the mass of the body and $\mt I$ is the inertia tensor, which is a diagonal matrix in the principal frame. For a three-dimensional system consisting of $N$ non-collinear rigid bodies that interact via a potential $U(\vt r, \vt q)$, we define the Hamiltonian in terms of its generalized coordinates ($\vt r_i$ and $\vt q_i$) and conjugate momenta ($\vt p_i$ and $\vt \pi_i$), as follows:~\cite{Abreu_2017}

@@

The equations of motion are derived from the Hamiltonian Eq.\eqref{eq:H_NVE} by $\dot{\vt r}_i = \partial \mathcal{H} / \partial \vt p_i$, $\dot{\vt p}_i = -\partial \mathcal{H} / \partial \vt r_i$, $\dot{\vt q}_i = \partial \mathcal{H} / \partial \vt \pi_i$, and $\dot{\vt \pi}_i = -\partial \mathcal{H} / \partial \vt q_i$.\cite{Goldstein2002}. For a rigid body composed of $n_p$ individual particles, the derivative $\partial U/\partial \vt r$ is obtained by
\[
\diff{U}{\vt r_i} = -\vt F_i = -\sum_{j=1}^{n_p} {\vt F_j},
\]
where $\vt F_j$ is the force applied on particle $j$, expressed in the space-fixed frame of reference, and thus $\vt F_i$ is the resultant force on the body. As demonstrated in Abreu~\textit{et al.},\citep{Abreu_2017} $\partial{U}/\partial \vt q_i$ is given by
\[
\diff{U}{\vt q_i} = -2 \mt C_i \vt \tau_i = -2 \mt C_i \sum_{j=1}^{n_p} {\vt \delta_j} \times {\vt F_j},
\]
where $\vt \tau_i$ is the resultant torque exerted on the body, and ${\vt \delta}_j$ $=$ $\tr{\mt A} {\vt d}_j$ is the position of the particle $j$, both expressed in the space-fixed frame. After carrying out the required differentiations of Eq.\eqref{eq:H_NVE}, we obtain the following system of equations for the rigid body motion:
\begin{subequations}
\label{eq:EDO_system}
\begin{align}
&\dot{\vt r}_i = \frac{1}{m_i} \vt p_i \\
&\dot{\vt p}_i = \mt F_i \\
&\dot{\vt q}_i = \frac{1}{2} \mt B_i \vt \omega_i \label{eq:EDO_q} \\
&\dot{\vt \pi}_i = \frac{1}{2} \mt \Omega_i \vt \omega_i + 2 \mt C_i \vt \tau_i \label{eq:EDO_pi}
\end{align}
\end{subequations}
with $\mt \Omega_i = \mt B_i(\vt \pi_i)$.

@@

In this context, the Hamiltonian of Eq.\eqref{eq:H_NVE} can be rewritten as
\begin{equation}
\label{eq:H_split_omega}
\mathcal{H} =  U(\vt r, \vt q) +  \sum_{i=1}^N \left( \frac{\tr{\vt p_i} \vt p_i}{2m_i} + \frac{1}{2} \sum_{k=1}^3 I_{ik} \omega_{ik}^2\right),
\end{equation}
from which we recognize partial Hamiltonians, each of one generates equations of motion with known analytical solutions. In particular, we have $\mathcal{H}_b = U(\vt r, \vt q)$, $\mathcal{H}_{t} = \frac{1}{2m}\tr{\vt p} \vt p$, and $\mathcal{H}_k = (1/2) I_k \omega_k^2$. Of corse, $\mathcal{H} = \mathcal{H}_b + \mathcal{H}_t + \sum_{k=1}^3 \mathcal{H}_k$. The NVE dynamics of this system is governed by a Liouville operator given by\cite{Abreu_2017}
\begin{equation}
\label{eq:full_operator}
i\!L_\text{NVE} = i\!L_t + i\!L_b + \sum_{k=1}^3 i\!L_k,
\end{equation}
where the individual operators are defined as follows:
\[
\begin{split}
&boost:\\
&i\!L_b = \sum_{i=1}^N \left( \tr{\vt F}_i \diff{}{\vt p_i} + 2 \tr{\vt \tau}_i \tr{\mt C}_i \diff{}{\vt \pi_i} \right) \\
&translation: \\
& i\!L_t = \sum_{i=1}^N \frac{\tr{\vt p}_i}{m_i}\diff{}{\vt r_i} \\
&rotation \; about \; a \; principal \; axis:\\
& i\!L_k = \frac{1}{2} \sum_{i=1}^N {\omega_i}_k \left( \tr{\vt q}_i\tr{\hat{\mt B}_k} \diff{}{\vt q_i} + \tr{\vt \pi_i}\tr{\hat{\mt B}_k} \diff{}{\vt \pi_i} \right)
\end{split}
\]
The individual operators can be evaluated analytically according to
\[
\begin{split}
&boost:\\
&{\vt p} = {\vt p}_0 + {\vt F}_0 t \\
&{\vt \pi} = {\vt \pi}_0 +  2 {\mt C}_0 {\vt \tau}_0 t\\
&translation:\\
&{\vt r} = {\vt r}_0 + (t/m) {\vt p}_0\\
&rotation \; about \; a \; principal \; axis:\\
&{\vt q}(t) = \cos\left(\frac{\omega_k t}{2}\right) \vt q_0 + \sin\left(\frac{\omega_k t}{2}\right) \hat{\mt B}_k \vt q_0 \\
&{\vt \pi}(t) = \cos\left(\frac{\omega_k t}{2}\right) \vt \pi_0 + \sin\left(\frac{\omega_k t}{2}\right) \hat{\mt B}_k \vt \pi_0
\end{split}
\]
Following Miller \textit{et al}.,\cite{Miller2002} each complete rotation is factorized as
\begin{align*}
\label{eq:splitting_rot}
e^{\Delta t i\!L_r} =  e^{\frac{\Delta t}{2} i\!L_3} e^{\frac{\Delta t}{2} i\!L_2} e^{\Delta t i\!L_1} e^{\frac{\Delta t}{2} i\!L_2} e^{\frac{\Delta t}{2} i\!L_3},
\end{align*}
where ${\Delta t}$ is the time step size. The splitting of the Liouville operator of Eq.~\eqref{eq:full_operator} we employ in this paper is
\begin{equation}
\label{eq:trotter_splitting_NVE}
e^{\Delta t i\!L_\text{NVE}} = e^{\frac{\Delta t}{2} i\!L_b} e^{\Delta t i\!L_r} e^{\Delta t i\!L_t} e^{\frac{\Delta t}{2} i\!L_b}.
\end{equation}

@@

We employ the technique of Sergi and Ferrario\cite{Sergi2001} to couple a single Nos\'e-Hoover chain\cite{Martyna1992} (NHC) composed of $M$ thermostats to a system of $N$ rigid bodies. This is simpler than the approach of Kamberaj \textit{et al}.,\cite{Kamberaj2005} who employed two thermostat chains independently coupled to the translational and rotational degrees of freedom. To this end, we consider extra coordinates $\eta_j$ and their conjugate momenta $p_{\eta_j}$, for $j$ ranging from $1$ to $M$. The conserved ``extended Hamiltonian'' in this case is
\begin{equation}
\label{eq:H_nvt}
H = \mathcal{H}(\vt r,\vt p,\vt q,\vt \pi) + \sum_{j=1}^{M}\frac{p_{\eta_j}^2}{2Q_j} + L k_b T\eta_1 + k_b T\sum_{j=2}^M \eta_j,
\end{equation}
where $\mathcal H$ is the same as in Eq.~\eqref{eq:H_NVE}, $k_b T$ is the Boltzmann constant multiplied by the system temperature, $L$ is a constant to be determined, and each $Q_j$ is an inertial parameter of thermostat $j$. As recommended in Ref.~\onlinecite{Martyna1992}, one can make $Q_1 = L k_b T t_d^2$ and $Q_j = k_b T t_d^2$ for $j \geq 2$, where $t_d$ is a characteristic time scale of the system. The equations of motion regarding generalized coordinates are identical to those of a Hamiltonian system, i.e. $\dot{\vt r}_i = {\partial H}/{\partial \vt p_i}$ and $\dot{\vt q}_i = {\partial H}/{\partial \vt \pi_i}$ for every particle $i$ and $\dot{\eta}_j = {\partial H}/{\partial {p_\eta}_j}$ for every thermostat $j$. However, the equations of motion for generalized momenta contain non-Hamiltonian terms. For instance, the first thermostat of the chain acts on the translational and rotational momenta of each particle $i$ as
\[
\begin{split}
\dot{\vt p}_i &= -\diff{H}{\vt r_i} - {\vt p}_i \diff{H}{p_{\eta_1}} \quad \text{and} \\
\dot{\vt \pi}_i &= -\diff{H}{\vt q_i} - {\vt \pi}_i \diff{H}{p_{\eta_1}}.
\end{split}
\]

@@

Finally, analysis of the probability distribution generated by the canonical equations of motion can be carried out via the generalized phase-space analysis of Tuckerman \textit{et al}.\cite{Tuckerman2001} For this, conservation laws must be identified and driven variables must be eliminated. For a system without external forces, the extended energy $H$ and the vector quantity $e^{\eta_1}\sum_{i=1}^N {\vt p}_i = \vt K$ are integrals of motion, while the center-of-mass position can be considered as a driven variable.\cite{Tuckerman2001} Moreover, $2N$ equations must be eliminated from the system due to the constraints involving $\vt q_i$ and $\vt \pi_i$ for all $i$. In the general case, by taking all these facts into account and applying the analysis of Ref.~\onlinecite{Tuckerman2001}, we can deduce that the metric determinant factor of the effective phase space is $\sqrt{g} = e^{(6N-2) \eta_1 + \sum_{j=2}^M \eta_j}$ and that the correct canonical distribution is attained if one makes $L = 6N$. In the particular (and rather common) situation in which one sets $\vt P = \vt 0$ at the beginning of a simulation, one must make $L = 6N - 3$ because all three elements of $\vt P$ will be conserved.\cite{Martyna1994}

@@

In turn, the propagator $e^{t i\!L_\text{NHC}^{(j)}}$ defined in Eq.~\eqref{eq:iL_NHC_j} behaves according to the equations $\dot{p}_{\eta_j} = G_j - \alpha_{j+1} p_{\eta_j}$ and $\dot{\eta}_{j+1} = \alpha_{j+1}$, with all variables other than $p_{\eta_j}$ and $\eta_{j+1}$ kept unchanged, whose analytical solution is
\begin{align*}
&p_{\eta_j}(t) = p_{\eta_j}^0 + \Big( G_j - \alpha_{j+1} p_{\eta_j}^0 \Big) \phi\left(\alpha_{j+1} t\right) t \\
&\eta_{j+1}(t) = \eta_{j+1}^0 + \alpha_{j+1} t
\end{align*}
where $\phi(x) = (1-e^{-x})/x$. We might note that, in order to avoid numerical issues with $x \rightarrow 0$, we evaluate $\phi(x)$ as $\sum_{n=0}^3 {(-1)^n x^n}/{(n+1)!}$ whenever $|x| \leq 10^{-4}$. The solution above is equivalent to that found in Ref.~\onlinecite{Martyna1996}.

@@

\begin{table}
	\caption{New results involving Kamberaj \text{et al}.\cite{Kamberaj2005}}
	\label{table:new_results}
	\begin{ruledtabular}
		\begin{tabular}{cccc}
			& \multicolumn{3}{c}{ Temperature (K) } \\
			\cline{2-4}
			$\Delta t$/fs & Proposed & Alternative & Kamberaj \text{et al}. \\
			\hline
			1.0 & $300.01 \pm 0.02$ & $300.00 \pm 0.02$ & $299.52 \pm 0.02$ \\
			2.0 & $300.01 \pm 0.03$ & $299.99 \pm 0.03$ & $298.24 \pm 0.03$ \\
			3.0 & $299.97 \pm 0.04$ & $299.98 \pm 0.04$ & $295.94 \pm 0.04$
		\end{tabular}
	\end{ruledtabular}
\end{table}

>>>>

%\documentclass[journal=jctcce,manuscript=article]{achemso}
\documentclass[journal=jctcce,manuscript=article,layout=twocolumn]{achemso}

% NOTE: comment bellow before submitting
\let\titlefont\undefined
\makeatletter
\let\l@addto@macro\relax
\makeatother
\usepackage[fontsize=11pt]{scrextend}
% NOTE: comment above before submitting

\usepackage{threeparttable}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage[T1]{fontenc}
\usepackage[table]{xcolor}
\definecolor{lightgray}{gray}{0.9}

\newcommand{\mt}[1]{\boldsymbol{\mathbf{#1}}}   % matrix symbol
\newcommand{\vt}[1]{\boldsymbol{\mathbf{#1}}}   % vector symbol
\newcommand{\tr}[1]{#1^\text{t}}                % transposition
\newcommand{\diff}[2]{\frac{\partial #1}{\partial #2}} % derivative
%\newcommand{\diff}[2]{\nabla_{#2}#1}            % derivative
\newcommand{\sdiff}[2]{\nabla^2_{#2}#1}         % 2nd derivative
\newcommand{\Ham}[1]{{\mathcal H}_\text{#1}}    % Hamiltonian
\newcommand{\Liu}[1]{i\!L_\text{#1}}            % Liouville

\usepackage{array}
\newcolumntype{L}{>{$}l<{$}}
\newcolumntype{C}{>{$}c<{$}}
\newcolumntype{R}{>{$}r<{$}}

\newenvironment{smallarray}[1]                          % small arrays
{\null\,\vcenter\bgroup\scriptsize
	\renewcommand{\arraystretch}{1.5}
	\arraycolsep=.13885em
	\hbox\bgroup$\left[\array{@{}#1@{}}}
{\endarray\right]$\egroup\egroup\,\null}
\listfiles

@@

\title{Revisiting the Nos\'{e}-Hoover chain thermostat for Rigid-Body Molecular Dynamics}

%\abbreviations{AAA, BBB, CCC}
\keywords{aaa, bbb, ccc}

\begin{tocentry}
	Graphical Abstract
\end{tocentry}

\begin{abstract}
	Abstract.
\end{abstract}

@@

Molecular Simulation, being the computational realization of Statistical Mechanics,\cite{Tuckerman2010} involves several types of approximations in order to mimic ``real'' physical systems. Naively, one would expect those approximations to only produce quantitative deviation from experimental values, and usually we tend to attribute any disagreement to a lack of accuracy of the force field. However, what can be hidden in the deviations are actually systematic errors, which may introduce non-physical behaviour. For instance, it has already been shown that the unavoidable truncation errors, associated with the numerical integration in Molecular Dynamics (MD), disrupt the equipartition of kinetic energy and, as a consequence, the temperature becomes an undefined property.\cite{Eastwood_2010} In a previous paper,\citep{Silveira_2017} we reported the same artifact to occur on the partition of kinetic energy between translational and rotational degrees of freedom in rigid-body MD. That finding brings into question the method introduced by Kamberaj and co-workers\cite{Kamberaj2005} which, assuming equipartition, couples separate chains of Nos\'{e}-Hoover thermostats on the rotational and translational degrees of freedom, but fails to reproduce the specified temperature for certain time step sizes.

@@

In this paper, we derive time reversible and measure-preserving numerical integrators for rigid-body MD in the canonical (NVT) ensemble. The starting point is our previous work on the microcanonical (NVE) ensemble,\citep{Silveira_2017} which employs the unit quaternion representation for the rotational degrees of freedom. Simultaneosuly coupling the translational and rotational degrees of freedom to a unique Nos\'{e}-Hoover chain thermostat, we are able to mantain the temperature of the system at its specified value. The analysis of the integrators includes comparisons with results from the method introduced by Kamberaj~\textit{et al.},\cite{Kamberaj2005} as well as with the Hybrid Monte Carlo (HMC) method,\cite{Duane1987} which is exact in the sense that truncation errors do not affect its accuracy. We also include results from simulations using the conventional SHAKE\cite{Ryckaert1977} procedure, also coupled to a unique Nos\'{e}-Hoover chain.

@@

Guided by a particular factorization of the rotation matrix expressed in terms of quaternion components, in a previous paper \cite{Silveira_2017} we derived symplectic numerical integrators for rigid bodies in the NVE ensemble. The first one uses an exact solution for torque-free rotations. The second approach is based on the method introduced by \citeauthor{Dullweber_1997} \cite{Dullweber_1997} and \citeauthor{Miller2002} \cite{Miller2002}, that employs an approximate solution for rotation. For mathematical details the reader is referred to our paper.\cite{Silveira_2017} Next, we restrict ourselves to present the expressions required to develop the NVT case.

@@

The matrix $\mt A$ which can be computed from a given unit quaternion $\vt q$ as explained in Refs.~\citenum{Allen1989,Miller2002} admits the following factorization:
\begin{equation}
\label{eq:factorization_of_A}
{\mt A} = \tr{\mt B}{\mt C},
\end{equation} 
where
\begin{equation}
\label{eq:def_B_and_C}
\mt B(\vt q) = \begin{smallarray}{rrrr}
-q_2 & -q_3 & -q_4 \\
 q_1 & -q_4 &  q_3 \\
 q_4 &  q_1 & -q_2 \\
-q_3 &  q_2 &  q_1
\end{smallarray}
\; \text{and} \;
\mt C(\vt q) = \begin{smallarray}{rrrr}
-q_2 & -q_3 & -q_4 \\
 q_1 &  q_4 & -q_3 \\
-q_4 &  q_1 &  q_2 \\
 q_3 & -q_2 &  q_1
\end{smallarray}.
\end{equation}
That factorization of matrix $\mt A$ provides the mathematical relation between the three-dimensional angular velocity vector $\vt \omega$ and the four-dimensional rate of change of the quaternion components $\dot{\vt q}$, which is given by
\begin{equation}
\label{eq:relation_qdot_omega}
\dot{\vt q} = \frac{1}{2} \mt B \vt \omega.
\end{equation}
For the sake of brevity, the argument of a matrix-valued function (like $\mt A$, $\vt B$, or $\mt C$) might be omitted, when such argument is $\vt q$.

In the Hamiltonian approach, we employ the linear momentum $\vt p = m \dot{\vt r}$, conjugated to $\vt r$, as well as the momentum $\vt \pi= 4 \mt B \mt I \tr{\mt B} \dot{\vt q} = 2 \mt B \mt I \vt \omega$, conjugated to $\vt q$\cite{Goldstein2002}. In these equations, $m$ is the mass of the body and $\mt I$ is the inertia tensor, which is a diagonal matrix in the principal frame. For a three-dimensional system consisting of $N$ non-collinear rigid bodies that interact via a potential $U(\vt r, \vt q)$, we define the Hamiltonian in terms of its generalized coordinates ($\vt r_i$ and $\vt q_i$) and conjugate momenta ($\vt p_i$ and $\vt \pi_i$), as follows:~\cite{Silveira_2017}

@@

The equations of motion are derived from the Hamiltonian Eq.\eqref{eq:H_NVE} by $\dot{\vt r}_i = \partial \mathcal{H} / \partial \vt p_i$, $\dot{\vt p}_i = -\partial \mathcal{H} / \partial \vt r_i$, $\dot{\vt q}_i = \partial \mathcal{H} / \partial \vt \pi_i$, and $\dot{\vt \pi}_i = -\partial \mathcal{H} / \partial \vt q_i$.\cite{Goldstein2002}. For a rigid body composed of $n_p$ individual particles, the derivative $\partial U/\partial \vt r$ is obtained by
\[
\diff{U}{\vt r_i} = -\vt F_i = -\sum_{j=1}^{n_p} {\vt F_j},
\]
where $\vt F_j$ is the force applied on particle $j$, expressed in the space-fixed frame of reference, and thus $\vt F_i$ is the resultant force on the body. As demonstrated in Abreu~\textit{et al.},\citep{Silveira_2017} $\partial{U}/\partial \vt q_i$ is given by
\[
\diff{U}{\vt q_i} = -2 \mt C_i \vt \tau_i = -2 \mt C_i \sum_{j=1}^{n_p} {\vt \delta_j} \times {\vt F_j},
\]
where $\vt \tau_i$ is the resultant torque exerted on the body, and ${\vt \delta}_j$ $=$ $\tr{\mt A} {\vt d}_j$ is the position of the particle $j$, both expressed in the space-fixed frame. After carrying out the required differentiations of Eq.\eqref{eq:H_NVE}, we obtain the following system of equations for the rigid body motion:
\begin{subequations}
\label{eq:EDO_system}
\begin{align}
&\dot{\vt r}_i = \frac{1}{m_i} \vt p_i \\
&\dot{\vt p}_i = \mt F_i \\
&\dot{\vt q}_i = \frac{1}{2} \mt B_i \vt \omega_i \label{eq:EDO_q} \\
&\dot{\vt \pi}_i = \frac{1}{2} \mt \Omega_i \vt \omega_i + 2 \mt C_i \vt \tau_i \label{eq:EDO_pi}
\end{align}
\end{subequations}
with $\mt \Omega_i = \mt B_i(\vt \pi_i)$.

@@

In this context, the Hamiltonian of Eq.\eqref{eq:H_NVE} can be rewritten as
\begin{equation}
\label{eq:H_split_omega}
\mathcal{H} =  U(\vt r, \vt q) +  \sum_{i=1}^N \left( \frac{\tr{\vt p_i} \vt p_i}{2m_i} + \frac{1}{2} \sum_{k=1}^3 I_{ik} \omega_{ik}^2\right),
\end{equation}
from which we recognize partial Hamiltonians, each of one generates equations of motion with known analytical solutions. In particular, we have $\mathcal{H}_b = U(\vt r, \vt q)$, $\mathcal{H}_{t} = \frac{1}{2m}\tr{\vt p} \vt p$, and $\mathcal{H}_k = (1/2) I_k \omega_k^2$. Of corse, $\mathcal{H} = \mathcal{H}_b + \mathcal{H}_t + \sum_{k=1}^3 \mathcal{H}_k$. The NVE dynamics of this system is governed by a Liouville operator given by\cite{Silveira_2017}
\begin{equation}
\label{eq:full_operator}
i\!L_\text{NVE} = i\!L_t + i\!L_b + \sum_{k=1}^3 i\!L_k,
\end{equation}
where the individual operators are defined as follows:
\[
\begin{split}
&boost:\\
&i\!L_b = \sum_{i=1}^N \left( \tr{\vt F}_i \diff{}{\vt p_i} + 2 \tr{\vt \tau}_i \tr{\mt C}_i \diff{}{\vt \pi_i} \right) \\
&translation: \\
& i\!L_t = \sum_{i=1}^N \frac{\tr{\vt p}_i}{m_i}\diff{}{\vt r_i} \\
&rotation \; about \; a \; principal \; axis:\\
& i\!L_k = \frac{1}{2} \sum_{i=1}^N {\omega_i}_k \left( \tr{\vt q}_i\tr{\hat{\mt B}_k} \diff{}{\vt q_i} + \tr{\vt \pi_i}\tr{\hat{\mt B}_k} \diff{}{\vt \pi_i} \right)
\end{split}
\]
The individual operators can be evaluated analytically according to
\[
\begin{split}
&boost:\\
&{\vt p} = {\vt p}_0 + {\vt F}_0 t \\
&{\vt \pi} = {\vt \pi}_0 +  2 {\mt C}_0 {\vt \tau}_0 t\\
&translation:\\
&{\vt r} = {\vt r}_0 + (t/m) {\vt p}_0\\
&rotation \; about \; a \; principal \; axis:\\
&{\vt q}(t) = \cos\left(\frac{\omega_k t}{2}\right) \vt q_0 + \sin\left(\frac{\omega_k t}{2}\right) \hat{\mt B}_k \vt q_0 \\
&{\vt \pi}(t) = \cos\left(\frac{\omega_k t}{2}\right) \vt \pi_0 + \sin\left(\frac{\omega_k t}{2}\right) \hat{\mt B}_k \vt \pi_0
\end{split}
\]
Following Miller \textit{et al}.,\cite{Miller2002} each complete rotation is factorized as
\begin{align*}
\label{eq:splitting_rot}
e^{\Delta t i\!L_r} =  e^{\frac{\Delta t}{2} i\!L_3} e^{\frac{\Delta t}{2} i\!L_2} e^{\Delta t i\!L_1} e^{\frac{\Delta t}{2} i\!L_2} e^{\frac{\Delta t}{2} i\!L_3},
\end{align*}
where ${\Delta t}$ is the time step size. The splitting of the Liouville operator of Eq.~\eqref{eq:full_operator} we employ in this paper is
\begin{equation}
\label{eq:trotter_splitting_NVE}
e^{\Delta t i\!L_\text{NVE}} = e^{\frac{\Delta t}{2} i\!L_b} e^{\Delta t i\!L_r} e^{\Delta t i\!L_t} e^{\frac{\Delta t}{2} i\!L_b}.
\end{equation}

@@

We employ the technique of Sergi and Ferrario\cite{Sergi2001} to couple a single Nos\'e-Hoover chain\cite{Martyna1992} (NHC) composed of $M$ thermostats to a system of $N$ rigid bodies. This is simpler than the approach of Kamberaj \textit{et al}.,\cite{Kamberaj2005} who employed two thermostat chains independently coupled to the translational and rotational degrees of freedom. To this end, we consider extra coordinates $\eta_j$ and their conjugate momenta $p_{\eta_j}$, for $j$ ranging from $1$ to $M$. The conserved ``extended Hamiltonian'' in this case is
\begin{equation}
\label{eq:H_nvt}
H = \mathcal{H}(\vt r,\vt p,\vt q,\vt \pi) + \sum_{j=1}^{M}\frac{p_{\eta_j}^2}{2Q_j} + L k_b T\eta_1 + k_b T\sum_{j=2}^M \eta_j,
\end{equation}
where $\mathcal H$ is the same as in Eq.~\eqref{eq:H_NVE}, $k_b T$ is the Boltzmann constant multiplied by the system temperature, $L$ is a constant to be determined, and each $Q_j$ is an inertial parameter of thermostat $j$. As recommended in Ref.~\citenum{Martyna1992}, one can make $Q_1 = L k_b T t_d^2$ and $Q_j = k_b T t_d^2$ for $j \geq 2$, where $t_d$ is a characteristic time scale of the system. The equations of motion regarding generalized coordinates are identical to those of a Hamiltonian system, i.e. $\dot{\vt r}_i = {\partial H}/{\partial \vt p_i}$ and $\dot{\vt q}_i = {\partial H}/{\partial \vt \pi_i}$ for every particle $i$ and $\dot{\eta}_j = {\partial H}/{\partial {p_\eta}_j}$ for every thermostat $j$. However, the equations of motion for generalized momenta contain non-Hamiltonian terms. For instance, the first thermostat of the chain acts on the translational and rotational momenta of each particle $i$ as
\[
\begin{split}
\dot{\vt p}_i &= -\diff{H}{\vt r_i} - {\vt p}_i \diff{H}{p_{\eta_1}} \quad \text{and} \\
\dot{\vt \pi}_i &= -\diff{H}{\vt q_i} - {\vt \pi}_i \diff{H}{p_{\eta_1}}.
\end{split}
\]

@@

Finally, analysis of the probability distribution generated by the canonical equations of motion can be carried out via the generalized phase-space analysis of Tuckerman \textit{et al}.\cite{Tuckerman2001} For this, conservation laws must be identified and driven variables must be eliminated. For a system without external forces, the extended energy $H$ and the vector quantity $e^{\eta_1}\sum_{i=1}^N {\vt p}_i = \vt K$ are integrals of motion, while the center-of-mass position can be considered as a driven variable.\cite{Tuckerman2001} Moreover, $2N$ equations must be eliminated from the system due to the constraints involving $\vt q_i$ and $\vt \pi_i$ for all $i$. In the general case, by taking all these facts into account and applying the analysis of Ref.~\citenum{Tuckerman2001}, we can deduce that the metric determinant factor of the effective phase space is $\sqrt{g} = e^{(6N-2) \eta_1 + \sum_{j=2}^M \eta_j}$ and that the correct canonical distribution is attained if one makes $L = 6N$. In the particular (and rather common) situation in which one sets $\vt P = \vt 0$ at the beginning of a simulation, one must make $L = 6N - 3$ because all three elements of $\vt P$ will be conserved.\cite{Martyna1994}

@@

In turn, the propagator $e^{t i\!L_\text{NHC}^{(j)}}$ defined in Eq.~\eqref{eq:iL_NHC_j} behaves according to the equations $\dot{p}_{\eta_j} = G_j - \alpha_{j+1} p_{\eta_j}$ and $\dot{\eta}_{j+1} = \alpha_{j+1}$, with all variables other than $p_{\eta_j}$ and $\eta_{j+1}$ kept unchanged, whose analytical solution is
\begin{align*}
&p_{\eta_j}(t) = p_{\eta_j}^0 + \Big( G_j - \alpha_{j+1} p_{\eta_j}^0 \Big) \phi\left(\alpha_{j+1} t\right) t \\
&\eta_{j+1}(t) = \eta_{j+1}^0 + \alpha_{j+1} t
\end{align*}
where $\phi(x) = (1-e^{-x})/x$. We might note that, in order to avoid numerical issues with $x \rightarrow 0$, we evaluate $\phi(x)$ as $\sum_{n=0}^3 {(-1)^n x^n}/{(n+1)!}$ whenever $|x| \leq 10^{-4}$. The solution above is equivalent to that found in Ref.~\citenum{Martyna1996}.

@@

\begin{table}
	\caption{New results involving Kamberaj \text{et al}.\cite{Kamberaj2005}}
	\label{table:new_results}
%	\begin{ruledtabular}
		\begin{tabular}{cccc}
			& \multicolumn{3}{c}{ Temperature (K) } \\
			\cline{2-4}
			$\Delta t$/fs & Proposed & Alternative & Kamberaj \text{et al}. \\
			\hline
			1.0 & $300.01 \pm 0.02$ & $300.00 \pm 0.02$ & $299.52 \pm 0.02$ \\
			2.0 & $300.01 \pm 0.03$ & $299.99 \pm 0.03$ & $298.24 \pm 0.03$ \\
			3.0 & $299.97 \pm 0.04$ & $299.98 \pm 0.04$ & $295.94 \pm 0.04$
		\end{tabular}
%	\end{ruledtabular}
\end{table}
